Put Your Money Where Your Mouth Is: Prediction Markets and Climate Change

### 2016-03-17 paper.Rmd
paper: abstract and other edits throughout.

---
title: 'Put Your Money Where Your Mouth Is: Prediction Markets and Climate Change'
author: "John J. Nay, Martin Van der Linden, Jonathan Gilligan"
date: "March 12, 2016"
output:
  pdf_document:
    keep_tex: yes
  html_document: default
bibliography: library.bib
---
```{r fig_ref, echo=FALSE}
source("utils/fig_ref.R") # tabRef and figRef functions
source("utils/multi_plot.R")
``` 
```{r, caption_setting, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(knitr)
knit_hooks$set(htmlcap = function(before, options, envir) {
  if(!before) {
    paste('<p class="caption">',options$htmlcap,"</p>",sep="")
    }
    })
# small margins
knit_hooks$set(small_mar = function(before, options, envir) {
    if (before) par(mar = rep(0.0001, 4))  # smaller margins all around
})
# opts_chunk$set(dev="tiff", 
#                dev.args=list(compression="lzw", type="cairo"),
#                dpi=300)
```
```{r multi_plot, echo=FALSE}
source("utils/multi_plot.R")
library(ggplot2)
theme_set(theme_classic())
library(pander)
```

@@

We present a simulation model as a computational test-bed for a climate prediction market. Traders adapt their beliefs about future global temperatures based on the monetary performance of their neighboring traders in a social network. We conduct experiments with the model to determine how a variety of market factors and the climate data-generating process may affect the extent to which climate prediction markets foster a convergence of market participants' beliefs regarding a true model of the global climate. We simulate two alternative climate futures: one where the natural log of CO$_2$ is the primary driver of global temperature and one where sun spots are the primary driver. These represent the two most plausible competing views in the public discourse. Market participation causes traders to converge toward believing the "true" climate model.

@@

Traders bet on global temperature anomalies six years in the future. 
During the six-year period, traders buy and sell futures. 
Every year, traders update their models and forecast of future temperatures based on newly available data. 
At the end of each six-year period, winners collect gains and traders revise beliefs about climate models, based on their ideology and the beliefs of top earners in their social network. 
In this section, we describe the models used to generate future temperature data, agents beliefs about those models, the market procedures, and the social network connecting agents. 
We conclude with details on model dynamics and an overview of the model parameters.

@@

An example of such social network is depicted in Figure X. 
We assume that the initial network is segregated.
Traders who initially believe in model 1 (resp. 2) are more likely to be linked with other traders who also believe in 1 (resp. 2). 
Although traders can change their approximate model over time, the connections between traders do not change as the market unfolds (i.e the edges are fixed).

@@

 At each time $t$, traders :

@@

The model depends on the following parameters, which we vary in simulation experiments to determine their effects on the convergence of beliefs.

@@

As a validation that the market is operating correctly, we run the market with actual temperatures 1880 to 2015 and market betting from 1931 to 2014. As warming signal begins to show up in 1980s, traders begin to converge toward the log(CO$_2$) model (`r figRef("time")`).

@@

The sensitivity analysis is based on a Latin Hypercube Sampling of parameter sets from the following distributions [@beachkofski_improved_2002; @carnell_lhs_2012].

@@

We use the model to simulate 10,000 outcomes based on the input paramter sets and then conduct a partial rank correlation coefficient analysis on the relationship between the input matrix, $X$, resulting simulated vector, $y$ [@Marino2008; @pujol_sensitivity:_2014; @saltelli_sensitivity_2009]. Partial correlation computes the linear relationship between the part of the variation of $X_i$ and $y$ that are linearly independent of other $X_j$  ($j \neq i$).  The only difference with the partial correlation and partial rank correlation (which we use here), is that the outcome variable $y$ is first ranked-transformed in order to allow us capture potentially non-linear relationships. That is every value, $y_s$, in the data set is replace by a number $f(y_s) \equiv |\{y_r~|~ y_s > y_r\}|$, where $y_r$ are the other observed values of $y$ in the data set.

Our sensitivity analysis, `r figRef("sa")`, shows that ideology, risk tolerance, the number of traders, the number of edges per trader, the segmentation of the social network, and the true model all significantly affect convergence. Segmented social networks can create an "echo-chamber" effect. Market participation causes traders to converge toward believing the "true" climate model; however, when the true model is CO$_2$-driven there is more convergence due to less noise in the relationship between predictor variable (log of CO$_2$ versus TSI) and temperature.

@@

```{r time, eval=FALSE, echo=FALSE, message=FALSE, htmlcap=figRef("time", "Convergence over trading sequences."), fig.width=10.5, fig.height=8.5}
source("utils/time_plot.R")
p1 <- history_box() + theme(text=element_text(family='Palatino'))
p2 <- future_box() + theme(text=element_text(family='Palatino'))
multi_plot(p1, p2, cols=2)
```

>>>>

---
title: 'Put Your Money Where Your Mouth Is: Prediction Markets and Climate Change'
author: "John J. Nay, Martin Van der Linden, Jonathan Gilligan"
date: "March 17, 2016"
output:
  pdf_document:
    keep_tex: yes
  html_document: default
bibliography: library.bib
---
```{r fig_ref, echo=FALSE}
source("utils/fig_ref.R") # tabRef and figRef functions
source("utils/multi_plot.R")
``` 
```{r, caption_setting, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(knitr)
knit_hooks$set(htmlcap = function(before, options, envir) {
  if(!before) {
    paste('<p class="caption">',options$htmlcap,"</p>",sep="")
    }
    })
# small margins
knit_hooks$set(small_mar = function(before, options, envir) {
    if (before) par(mar = rep(0.0001, 4))  # smaller margins all around
})
# opts_chunk$set(dev="tiff", 
#                dev.args=list(compression="lzw", type="cairo"),
#                dpi=300)
```
```{r multi_plot, echo=FALSE}
source("utils/multi_plot.R")
library(ggplot2)
theme_set(theme_classic())
library(pander)
```

@@

We present a simulation model as a computational test-bed for a climate prediction market. Traders adapt their beliefs about future global temperatures based on the monetary performance of their neighboring traders in a social network. We conduct experiments with the model to determine how a variety of market factors and the temperature data-generating process may affect the extent to which climate prediction markets foster a convergence of market participants' beliefs regarding a true model of the global climate. We simulate two alternative climate futures: one where the natural log of CO$_2$ is the primary driver of global temperature and one where sun spots are the primary driver. These represent the two most plausible competing views in the public discourse. Market participation causes traders to converge toward believing the "true" climate model under a variety of model parameterizations in a relatively small number of years, suggesting that a climate prediction market could be useful regardless of which climate change beliefs ultimately turn out to be the best model of temperature change.

@@

Traders bet on global temperature anomalies six years in the future. 
During the six-year period, traders buy and sell futures. 
Every year, traders update their models and forecast of future temperatures based on newly available data. 
At the end of each six-year period, winners collect gains and traders revise beliefs about climate models, based on their ideology and the beliefs of top earners in their social network. 
In this section, we describe the models used to generate future temperature data, agents beliefs about those models, the market procedures, and the social network connecting agents. 
We conclude with details on model dynamics and an overview of the model parameters we experimentally vary.

@@

An example of such social network is depicted in Figure X. 
We assume that the initial network is segregated.
Traders who initially believe in model 1 (resp. 2) are more likely to be linked with other traders who also believe in 1 (resp. 2). 
This reflects reality where cultural and social groups are correlated with climate change beliefs. 
Although traders can change their approximate model over time, the connections between traders do not change as the market unfolds, i.e the edges are fixed.

@@

 At each time $t$, traders:

@@

The model depends on the following parameters, which we vary in simulation experiments to determine their effects on the convergence of beliefs. We group the parameters into network and individual behavioral factors.

@@

As a validation that the market is operating correctly, we run the market with actual temperatures 1880 to 2015 and market betting from 1931 to 2014. As a warming signal begins to show up in 1980s, traders begin to converge toward the log(CO$_2$) model (`r figRef("time")`).

@@

Our primary focus is on simulating the market in future scenarios. This is more relevant for policy design and more interesting theoretically due to the increasing divergence of global temperature values under the two future scenarios. The sensitivity analysis is based on a Latin Hypercube Sampling of parameter sets from the following distributions [@beachkofski_improved_2002; @carnell_lhs_2012].

@@

We use the model to simulate 10,000 outcomes based on the input paramter sets and then conduct a partial rank correlation coefficient analysis on the relationship between the input matrix, $X$, and the resulting simulated outcome vector of belief convergence scores, $y$ [@Marino2008; @pujol_sensitivity:_2014; @saltelli_sensitivity_2009]. Partial correlation computes the linear relationship between the part of the variation of $X_i$ and $y$ that are linearly independent of other $X_j$  ($j \neq i$).  The only difference with the partial correlation and the partial rank correlation that we use here is that the outcome variable $y$ is first ranked-transformed in order to allow us capture potentially non-linear relationships. That is every value, $y_s$, in the data set is replace by a number $f(y_s) \equiv |\{y_r~|~ y_s > y_r\}|$, where $y_r$ are the other observed values of $y$ in the data set.

Our sensitivity analysis, `r figRef("sa")`, shows that ideology, risk tolerance, the number of traders, the number of edges per trader, the segmentation of the social network, and the true model all statistically significantly affect convergence. Segmented social networks can create an "echo-chamber" effect. Market participation causes traders to converge toward believing the "true" climate model; however, when the true model is CO$_2$-driven there is more convergence due to less noise in the relationship between predictor variable (log of CO$_2$ versus TSI) and temperature.

@@

```{r time, eval=TRUE, echo=FALSE, message=FALSE, htmlcap=figRef("time", "Convergence over trading sequences."), fig.width=10.5, fig.height=8.5}
source("utils/time_plot.R")
p1 <- history_box() + theme(text=element_text(family='Palatino'))
p2 <- future_box() + theme(text=element_text(family='Palatino'))
multi_plot(p1, p2, cols=2)
```
